Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects

### 2025-05-10 paper.qmd
final version after extension

This work focusses on the European region. The hourly flight density
estimation is based on on flown \acr{IFR} trajectories as recorded by
EUROCONTROL's \acr{NM}[^1]. To decrease spatial complexity, the data is
binned using Uber's H3 geospatial indexing system [@uber-h3]. The
reduction of the spatial complexity enables fast processing. For this
work, we employ a H3-cell-resolution of 3. This corresponds to hexagons
of an average area $\qty{12393}{km^2}$ and an average edge length of
$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.

<!-- CAN WE MOTIVATE WHY resolution 3??? -- e.g. is this an area used in other works, or does it do something? -->

@@

To build an upper threshold for the flight exposure, this study uses the
busiest week during the \acr{IATA} summer ISO week in 2023 and 2024.
This ensures a reasonable value, as peak airspace utilisation provides a handle on the air traffic operations and interactions of the different aircraft as part of wider *'traffic flows'*.
<!-- does ISO do here anything? --> The flight densities are
characterised on an hourly basis for all days in these weeks per H3-cell.

<!-- next sentence feels like an orphan ... do we need this here? ------>
This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and provides a reproducible example for States, Agencies or relevant authorities to build upon.

@@

<!-- next sentence feels like an orphan ... do we need this here? ------>
The next sections will develop the mathematical background for the
evaluation of the collision risk for which air traffic density is
a key component.

@fig-density-summer2024-res-3 shows the hourly flight density for the 5th of
July 2024. The chart depicts the expected outcomes in terms of major
traffic flows, and areas of high or low traffic. For example, cells with
a density of zero are clearly visible for areas of war or no-fly zones
(Ukraine, Libya or Syria). At the same time, the major traffic flows (North Atlantic,
Canaries and East-bound over Türkiye) are clearly highlighted showing
densities of more than 10.

@@

Flight density in the EUROCONTROL Area on the 5th of July 2024, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$). The black hexagon in the South of UK marks the densest cell at 24.56 flights/hour, just around the airport of London Heathrow.
:::

@@

Regarding the aircraft-debris encounter geometry, this yields to two conceptual considerations:

1. if the aircraft were stationary, the area exposed to the debris impact would be the top face of the rectangular solid, and
2. if the debris were stationary, it would be the front face.

@@

* $s_{a}$ is the average aircraft speed, 
* $s_{d_k}$ is the average fall speed of the $k$-th debris object, 
* $a_T$ area of the aircraft as viewed from the top, and 
* $a_F$ area of the aircraft as viewed from the from the front of the aircraft.

@@

@@

@@

Impact probability density $\sigma(\varphi, i = 51.7)$ for debris of
inclination $i$ and altitude $\qty{550}{km}$.
:::

The impact probability density $\sigma(\varphi, i)$ for an inclination $i$ of
$\ang{51.7}$ is plotted in @fig-weighting-inclination-i and takes a characteristic U shape with zero values for latitudes beyond $i$.
This density has been derived by simulating a full circular orbit as detailed
in @#sec-orbitals.

@@

::: {#fig-weighting}
```{r weighting}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/weighting.png")
```

@@

::: {#fig-weighting-nm-area}
```{r weighting-nm-area}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/weighting_on_Earth.png")
```

Hourly weightings for H3 cells, $W_H(h)$, in \acr{NM} area. 
:::

@@

@@

When combining the hourly ($H$) aircraft density (per aircraft type) with the
relevant weightings per $h$ cell, $W_H(h)$, we can numerically compute the
hourly impact probability for each hexagonal cell as:

@@

* $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
* $W(h)$ is the mean value of the weighting function over the area of hexagon
$h$,
* $\rho_{t,H}(h)$ is the traffic density for the \acr{ICAO} aircraft type $t$ in
cell $h$ at hour $H$, and
* $\varepsilon_t$ is effective exposed area for aircraft of type $t$.

@@

# Use Case: KOSMOS 482 Descent Craft re-entry
<!-- I would make this a separate section in the paper -->

In few days (at the time of writing of this paper) at 10:05 May 10, 2025
($\pm 1.5$ days), the KOSMOS 482 Descent Craft should re-enter the
Earth's atmosphere [@kosmos482reentry].
This is the lander module from a 1972 failed USSR Venera mission [@SpaceReviewKosmos],
it is reported to weigh about $\qty{480}{kg}$ and to have a diameter of approximately $\qty{1}{m}$.
Being built to survive Venus's atmosphere which is much denser and hotter than Earth's one (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
it could re-enter Earth intact and not break into smaller fragments some of which could burn up.

Knowing that the debris inclination is $\ang{51.7}$, see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a proxy for the traffic density on the day of re-entry.
Using the approach in this study and the weightings for this specific debris,
the cell with the highest impact risk is cell, `83194afffffffff` (located South of London), with a value of $1.89 \cdot  10^{-10}$.

@@

Eventually once the re-entry track is known, the H3 cells around \qty{35}{km} on
each side  can be identified to retrieve the values of collision expectancy.
It can then be decided whether it is advisable to close the airspace and re-route
traffic around the area affected.

@@

We have calculated collision risk probabilities for the two
busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
as logically depicted in @fig-data-densities.
The processing of data for 1 day takes less than 30 min from
data extraction to impact probability calculation on a
high-end laptop (MacBook Pro M1 Max).

::: {#fig-data-densities}
```{r densities}
#| out-width: "80%"
#| echo: false
knitr::include_graphics("media/figures/cubes.png")
```

@@

::: {#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="h"}
```{r ellipse-parameters}
#| fig-cap: "Keplerian elements: semi-major axis, $a$, the distance between the foci, $2c$, the eccentricity, $e = (2c)/(2a)$. (the semi-parameter (a.k.a. _semilatus rectum_), $p$, describes the size of the conic section by defining the width at the primary focus, and is defined as $p = b^2 / a = a (1 - e^2)$)"

Definition of Keplerian elements. The orbital elements describing the
trajectory of orbiting object are illustrated in association with an
inertial reference frame $\hat{I}$-$\hat{J}$-$\hat{K}$. The $I$-axis,
$\hat{I}$, points toward the vernal equinox and the $J$-axis, $\hat{J}$,
is perpendicular to the $I$-axis in the equatorial plane n the sense of
Earth's rotation with the $K$-axis pointing up. The semi-major axis,
$a$, the eccentricity, $e$, the inclination, $i$, the right ascension of
the ascending node, $\Omega$, and the argument of perigee, $\omega$,
constitute, together with the mean anomaly at the initial position,
$M_0$, a set of six constants of motion that are preserved in the
absence of perturbation. True anomaly, $\nu$, specifies the location of
an orbiting object within the orbit. It is the angle between perigee and
the orbiting object's position vector measured in the direction of the
orbiting object's motion. Figures from [@sellers2003understanding],
respectively Figure 5-2 and Figure 5-9.
:::

@@

The findings are applied to a use-case: the imminent re-entry of the
lander module of an earlier failed Venus mission. The maximum collision
expectation for the busiest day in 2024 is determined.

@@

The approach presented in this paper builds on previous research.
Our approach utilises H3-tessellation to reduce the spatial complexity.
This cell indexing approach reduces the computation significantly.

@@

# Appendix: Orbital Elements {#sec-orbitals}

@@

The angular position, $\theta$, of the re-entry point from the ascending
node[^ascending-node] in the orbit plane is $\theta = \omega + \nu$. It is related to the
latitude, $\varphi$, via the Cartesian coordinate $k$,
$k = R \sin(\varphi) = R \sin(\theta)\sin(i)$, or

@@

Mass of Earth, $M_{\oplus} = \qty{5.9722e24}{kg}$.

These are used to derive numerically compute the weighting function,
@fig-weighting-inclination-i:
for debris with inclination $i$ we simulate an orbit of $N$, i.e. $10000$,
positions and calculate the corresponding Cartesian position and the
equivalent longitude and latitude.

>>>>

@@

This work focusses on the European region. The hourly flight density
estimation is based on flown \acr{IFR} trajectories as recorded by
EUROCONTROL's \acr{NM}[^1]. To decrease spatial complexity, the data is
binned using Uber's H3 geospatial indexing system [@uber-h3]. The
reduction of the spatial complexity enables fast processing. For this
work, we employ a H3-cell-resolution of 3. This corresponds to hexagons
of an average area $\qty{12393}{km^2}$ and an average edge length of
$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.

The choice of hex resolution is a compromise between having to use
highly detailed trajectories (more 4D positions and more processing and
storage) and capturing long enough portions of flight inside the cell.

To build an upper threshold for the flight impact exposure, this study uses the
busiest week during summer 2023 and 2024.
This ensures a reasonable value, as peak airspace utilisation provides
a handle on the air traffic operations and interactions of the different
aircraft as part of wider *'traffic flows'*.
The flight densities are characterised on an hourly basis for all days
in these weeks per H3-cell.

@@

@@

Flight density in the EUROCONTROL Area on the 5th of July 2024 (the busiest day of the busiest summer week in 2024) at H3 cells resolution $3$. The black hexagon in the South of UK marks the densest cell at 24.56 flights/hour, just around the airport of London Heathrow.
:::

@fig-density-summer2024-res-3 shows an example of binning the hourly flight
density on the 5th of July 2024 using H3 hexagons.
The chart depicts the expected outcomes in terms of major
traffic flows, and areas of high or low traffic. For example, cells with
a density of zero are clearly visible for areas of war or no-fly zones
(Ukraine, Libya or Syria). At the same time, the major traffic flows (North Atlantic,
Canaries and East-bound over Türkiye) are clearly highlighted showing
densities of more than 10.

The next sections will develop the mathematical background for the
evaluation of the collision risk for which air traffic density is
a key component.

@@

Regarding the aircraft-debris encounter geometry, this yields to two conceptual considerations:
1. if the aircraft were stationary, the area exposed to the debris impact would be the top face of the rectangular solid, and
2. if the debris were stationary, it would be the front face.

@@

where $s_{a}$ is the average aircraft speed, 
$s_{d_k}$ is the average fall speed of the $k$-th debris object, 
$a_T$ area of the aircraft as viewed from the top, and 
$a_F$ area of the aircraft as viewed from the from the front of the aircraft.

@@

Impact probability density $\sigma(\varphi, i = 51.95)$ for debris of
inclination $i$ and altitude $\qty{550}{km}$.
:::

@@

::: {#fig-weighting}
```{r weighting}
#| out-width: "100%"
#| echo: false
#| fig-pos: H
knitr::include_graphics("media/figures/weighting.png")
```

@@

The impact probability density $\sigma(\varphi, i)$ for an inclination $i$ of
$\ang{51.95}$ is plotted in @fig-weighting-inclination-i and takes a characteristic U shape with zero values for latitudes beyond $i$.
This density has been derived by simulating a full circular orbit as detailed
in @sec-orbitals.

@@

@@

![Hourly weightings for H3 cells, $W_H(h)$, in \acr{NM} area.](media/figures/weighting_on_Earth.png){#fig-weighting-nm-area out-width=100%}

@@

## Traffic density per cell

In order to derive $\rho_t(h)$, we have taken the trajectories of flown flights
in the \acr{NM} Area, resampled at $\qty{30}{s}$ so that the calculation of the
occupancy, i.e. the time spent in the hex, per aircraft type is just a matter
of counting the number of points per cell.
This binning operation is extremely fast to perform in Uber's H3 gridding system
and a real advantage in processing the huge amount of data point representing
the flight trajectories.

For July, $5^{th}$ 2024 alone, we extracted $37721$ flights with
$33,482,686$ 4D positions (at $\qty{30}{s}$ sampling), where 19 aircraft types
cover each at least 1% of the total and the remaining 266 (!) types account
for the residual 20.8%, see @fig-aircraft-types.

@@

When combining the hourly ($H$) aircraft occupancy (per aircraft type) with the
relevant weightings per $h$ cell, $W_H(h)$, and the effective exposed
area, $\varepsilon_t$, we can numerically compute the
hourly impact probability for each hexagonal cell as:

@@

where $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$.

::: {#fig-aircraft-types}
```{r aircraft-types-july-2024}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/aircraft_types.png")
```

Aircraft types on 2024-07-05 in \acr{NM} Area. 
:::

@@

# Use Case: KOSMOS 482 Descent Craft re-entry

While this paper was being written, the KOSMOS 482 Descent Craft was being
monitored for an uncontrolled re-entry, forecast at 06:39 May 10, 2025
($\pm 1.5$ hours) [@kosmos482reentry].

![Venera-4 lander mock-up at the Memorial Museum of Cosmonautics. The Kosmos 482 Descent Craft is probably similar.](media/venera-capsule-in-museum.jpg){#fig-kosmos width=80%}

This is the lander module from a 1972 failed USSR Venera mission
[@SpaceReviewKosmos], which is reported to weigh about $\qty{480}{kg}$ and
to have a diameter of approximately $\qty{1}{m}$.
Being built to survive Venus's atmosphere which is much denser and hotter than
Earth's one (surface pressure 93 times that of Earth and temperatures reaching
\ang{467}C (\ang{872}F)), it could re-enter Earth intact and not break into
smaller fragments some of which could burn up.

Knowing that the debris inclination is $\ang{51.95}$, 
see @fig-weighting-inclination-i, we can derive the maximum
collision expectation using data for the busiest day in the busiest 2024 week
as a proxy for the traffic density on the day of re-entry.
Using the approach in this study and the weightings for this specific debris,
the cell with the highest impact risk is cell, `83194afffffffff` 
(located South of London), with a value of $1.89 \cdot  10^{-10}$.

@@

Radar systems over Germany saw the descent craft at approximately 04:30 UTC and
06:04 UTC on May $10^{th}$. Given that the descent craft was not spotted at
the next expected pass at 07:32 UTC, it is likely that the re-entry already occurred before.

In general case when the re-entry track for a debris would be known
with some lead time, the H3 cells around, say,  \qty{35}{km} on each side
could be identified to retrieve the values of collision expectancy.
It would then be possible to decide whether it would advisable to close
the airspace and re-route traffic around the affected area.

::: {#fig-data-densities}
```{r densities}
#| out-width: "70%"
#| echo: false
#| fig-pos: H
knitr::include_graphics("media/figures/cubes.png")
```

@@

We have calculated collision risk probabilities for the two
busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
as logically depicted in @fig-data-densities.

The processing of data for 1 day takes less than 30 min from
data extraction to impact probability calculation on a
high-end laptop (MacBook Pro M1 Max).

@@

@@

The findings are applied to a use-case: the imminent re-entry of the
lander module of an earlier failed Venus mission. The maximum collision
expectation is determined using the busiest day in 2024 as a representative
proxy of the traffic on the day of re-entry.

@@

The approach presented in this paper builds on previous research.
Our approach utilises H3-tessellation to reduce the spatial complexity.
This cell indexing approach reduces the computation burden significantly.

@@

::: {#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="h" height=80%}
```{r ellipse-parameters}
#| fig-cap: "Keplerian elements: semi-major axis, $a$, the distance between the foci, $2c$, the eccentricity, $e = (2c)/(2a)$. (the semi-parameter (a.k.a. _semilatus rectum_), $p$, describes the size of the conic section by defining the width at the primary focus, and is defined as $p = b^2 / a = a (1 - e^2)$)"

Orbital elements, Figures 5-2 and 5-9 from [@sellers2003understanding].
:::

# Appendix: Orbital Elements {#sec-orbitals .appendix}

@@

The orbital elements describing the trajectory of orbiting object are
illustrated in association with an inertial reference frame
$\hat{I}$-$\hat{J}$-$\hat{K}$. 
The $I$-axis, $\hat{I}$, points toward the vernal equinox and 
the $J$-axis, $\hat{J}$, is perpendicular to the $I$-axis in the equatorial plane
in the sense of Earth's rotation, with the $K$-axis pointing up.
The semi-major axis, $a$, the eccentricity, $e$, the inclination, $i$,
the right ascension of the ascending node, $\Omega$, and the argument of perigee, $\omega$, constitute, together with the mean anomaly at the initial position,
$M_0$, a set of six constants of motion that are preserved in the
absence of perturbation. True anomaly, $\nu$, specifies the location of
an orbiting object within the orbit. It is the angle between perigee and
the orbiting object's position vector measured in the direction of the
orbiting object's motion.

The angular position, $\theta$, of the re-entry debris taken from the ascending
node[^ascending-node] in the orbit plane is $\theta = \omega + \nu$.
It is related to the latitude, $\varphi$, via the Cartesian coordinate $k$,
$k = R \sin(\varphi) = R \sin(\theta)\sin(i)$, or

@@

In the derivation of debris orbital positions, we use the value of the mass of Earth, $M_{\oplus} = \qty{5.9722e24}{kg}$ but we neglect that of the debris
because many orders of magnitudes smaller (and not always known).

Of all the orbital elements, for the needs of this paper to numerically compute
the weighting function, , @fig-weighting-inclination-i, we only need
the inclination $i$. 
For debris with inclination $i$ we simulate an orbit of $N$, i.e. $10000$,
positions ($\nu$) and calculate the corresponding Cartesian positions and
the equivalent longitudes and latitudes.
The latitudes are then binned on half degree intervals to compute
$\sigma(\varphi, i)$, as per @eq-impact-probability-density.
